Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-14 aistats.tex
Remove frederik's figure

We now introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
\begin{align}
	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
\end{align}
which is the common notion of \textit{entropy} in information theory.
By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$
(see Fig.~\ref{fig:duality}), giving
\aligns{
	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
}
\begin{figure}[t]
	\centering
	\includegraphics{duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
	}
	\label{fig:duality}
\end{figure}%
%
%Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
{\bf The Bregman Divergence} induced by $\logpart$,
measures the discrepancy between two parameters $\nat$ and $\nat_0$,
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0)
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
    \label{eq:defBregman}
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$.
In general, Bregman divergences are not symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
%Finally, it is equal to the divergence of its convex conjugate with switched arguments,
%\begin{align}
%	\bregman (\nat ; \nat_0)
%    = \bregmanconj ( \meanp_0 ; \meanp) \; .
%    \label{eq:bregman_switch}
%\end{align}

@@

\begin{problem}[Bounding MAP and MLE]
Upper bound the following quantities:
\begin{align}
	\label{eq:bregmanMLE}
	\text{MLE: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
	\label{eq:bregmanMAP}
	\text{MAP: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
\end{align}
where the expectation is on the data $\mD = (T_1, \dots, T_n)$.
\end{problem}

@@

A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its density reads
\[
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}.
\]
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
\textcolor{red}{The explanation here is not clear enough, or too concise} Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
\begin{align}
	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
	\conj(\m) &= - \half \log(\m) + \half \log\frac{\pi}{2} - \half  \; .
\end{align}
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m_*; \m_n)
	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
	\bregman( \nat_n; \nat_* )
	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in \cref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptote~\eqref{eq:asymptote} .
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

>>>>

We now introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
\begin{align}
	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
\end{align}
which is the common notion of \textit{entropy} in information theory.
By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$, giving
\aligns{
	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
}
%\begin{figure}[t]
%	\centering
%	\includegraphics{duality}
%	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$. Figure copied from \citet{kunstner2020homeomorphic}. % TODO make an original one.
%	}
%	\label{fig:duality}
%\end{figure}%
%
%Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
{\bf The Bregman Divergence} induced by $\logpart$,
measures the discrepancy between two parameters $\nat$ and $\nat_0$,
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0)
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
    \label{eq:defBregman}
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$.
In general, Bregman divergences are not symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
%Finally, it is equal to the divergence of its convex conjugate with switched arguments,
%\begin{align}
%	\bregman (\nat ; \nat_0)
%    = \bregmanconj ( \meanp_0 ; \meanp) \; .
%    \label{eq:bregman_switch}
%\end{align}

@@

\begin{problem}[Upper-bounding MAP and MLE]
Upper bound the following quantities:
\begin{align}
	\label{eq:bregmanMLE}
	\text{MLE: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
	\label{eq:bregmanMAP}
	\text{MAP: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
\end{align}
where the expectation is on the data $\mD = (T_1, \dots, T_n)$.
\end{problem}

@@

A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its density reads
\[
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}.
\]
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
\textcolor{red}{The explanation here is not clear enough, or too concise} 
Now we can match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
\begin{align}
	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
	\conj(\m) &= - \half \log(\m) + \half \log\frac{\pi}{2} - \half  \; .
\end{align}
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m_*; \m_n)
	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
	\bregman( \nat_n; \nat_* )
	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in \cref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptote~\eqref{eq:asymptote} .
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.
